Split-Apply-Combine with Dynamic Grouping

### 2024-03-27 paper/article.Rnw
tables on page 2 are now in a figure environment

\usepackage{amsmath, amssymb}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{tikz}

% shortcut for surjection arrow, and surjection version of xrightarrow
\newcommand{\onto}{\twoheadrightarrow}
\newcommand\xonto[2][]{%
  \mathrel{\ooalign{$\xrightarrow[#1\mkern4mu]{#2\mkern4mu}$\cr%
  \hidewidth$\rightarrow\mkern4mu$}}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dynamic Grouping} \label{sect:dynamicgrouping}
In this section the illustrate the concept of dynamic grouping with a minimal
worked example. This example is not very realistic, but it is constructed
to be simple enough so the whole procedure can be followed in detail.

@@

The following tables illustrate the idea. The left table represents the data
set to be aggregated by $A$ and $B$.  The table on the right represents the
output. Colors indicate which data was used. 
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{Input data}\\
\hline
$A$ & $B$ & $B_1$ & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   3   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 5   \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 6   \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  7   \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  8   \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  9   \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{Output}\\
\hline
$A\times B$ & \code{level} & $\mu_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 2  \\
\rowcolor{green!25} 2 12        & 1             & 5  \\
\rowcolor{green!25} 2 13        & 1             & 5  \\
\rowcolor{blue!25}  3 21        & 2             & 8  \\
\rowcolor{blue!25}  3 22        & 2             & 8  \\
\rowcolor{blue!25}  3 12        & 2             & 8  \\
\hline
\end{tabular}
\label{eq:example}
\end{center}
%
The First row in the output represents group $(A=1,B=11)$. The collapsing level
is zero, which means that no collapsing was necessary. Indeed, in the data
table we see that there are three rows with $A=1$ and $B=11$ with $Y$ values
$1, 2$, and $3$, resulting in $\mu_Y=(1+2+3)/3=2$ for this group.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensive Example: Economic Data} \label{sect:example}
In this section we discuss three practical examples using a synthetic
dataset included with the package.
<<>>=
data("producers")
head(producers)
@
This \code{producers} dataset contains synthetic data records of various income
sources for \Sexpr{nrow(producers)} industrial produces. The records are
classified into a local version of the NACE classification called \code{sbi}
and into \code{size} classes with values in $\{5,6,7,8,9\}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formal Description and Algorithms} \label{sect:algorithm}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to demonstrate how the algorithm must be generalized to
allow for a collapsing scheme.

>>>>

\usepackage{amsmath, amssymb}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{tikz}
\usepackage{float}
% shortcut for surjection arrow, and surjection version of xrightarrow
\newcommand{\onto}{\twoheadrightarrow}
\newcommand\xonto[2][]{%
  \mathrel{\ooalign{$\xrightarrow[#1\mkern4mu]{#2\mkern4mu}$\cr%
  \hidewidth$\rightarrow\mkern4mu$}}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dynamic Grouping} \label{sect:dynamicgrouping}
In this Section the illustrate the concept of dynamic grouping with a minimal
worked example. This example is not very realistic, but it is constructed
to be simple enough so the whole procedure can be followed in detail.

@@

The tables in Figure~\ref{fig:example} illustrate the idea. The left table
represents the data set to be aggregated by $A$ and $B$.  The table on the
right represents the output. Colors indicate which data was used. 
%
\begin{figure}[H]
\centering
\begin{tabular}{cccl}
\multicolumn{4}{l}{Input data}\\
\hline
$A$ & $B$ & $B_1$ & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   3   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 5   \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 6   \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  7   \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  8   \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  9   \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{Output aggregates}\\
\hline
$A\times B$ & \code{level} & $\mu_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 2  \\
\rowcolor{green!25} 2 12        & 1             & 5  \\
\rowcolor{green!25} 2 13        & 1             & 5  \\
\rowcolor{blue!25}  3 21        & 2             & 8  \\
\rowcolor{blue!25}  3 22        & 2             & 8  \\
\rowcolor{blue!25}  3 12        & 2             & 8  \\
\hline
\end{tabular}
\label{fig:example}
\caption{Input data (left) with grouping variables $A$, $B$ and $B_1$, and
means of $Y$ per $A\times B$ after dynamic grouping (right).}
\end{figure}
%
The First row in the output represents group $(A=1,B=11)$. The collapsing level
is zero, which means that no collapsing was necessary. Indeed, in the data
table we see that there are three rows with $A=1$ and $B=11$ with $Y$ values
$1, 2$, and $3$, resulting in $\mu_Y=(1+2+3)/3=2$ for this group.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensive Example: Economic Data} \label{sect:example}
In this Section we discuss three practical examples using a synthetic
dataset included with the package.
<<>>=
data("producers")
head(producers)
@
This \code{producers} dataset contains synthetic data records of various income
sources for \Sexpr{nrow(producers)} industrial produces. The records are
classified into a local version of the NACE classification called \code{sbi}
and into \code{size} classes with values in $\{5,6,7,8,9\}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formal Description and Algorithms} \label{sect:algorithm}
In this Section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to demonstrate how the algorithm must be generalized to
allow for a collapsing scheme.
